---
title: "EDA"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
## Smoker epigenetic dataset
A CpG island or CpG site is a part of the DNA where the GC content is greater than 50%. In this dataset methylation values of certain CpG sites are displayed with also the age, gender and smoking status for 671 people.

## Research Question
Is it possible to identify a person's gender, age or status of smoking given their CpG values?

\newpage
# EDA
## Codebook
```{r}
myData <- read.csv("data/Smoker_Epigenetic_df.csv")
myData <- myData %>% drop_na()

columns = colnames(myData[1:4])
columns <- append(columns, "Columns 5-24")
names <- c("Sample Accessions numbers", "Smoking status", "Gender", "Age", "CG Island")
type <- c("chr", "chr", "chr", "int", "num")
unit <- c(NA, "current/never", "f/m", NA, NA)
descriptions = c("GSM identifier testsubject", "Wether the person is smoking or not", "Gender", "Age", "Methylation Rate of CG Island")
codebook <- data.frame(columns, names, type, descriptions)
write.csv(codebook, "Codebook.csv", row.names = FALSE)
knitr::kable(codebook)
```
\newpage

## Data exploration
### Visualization

```{r CpG-tables}
subsetdata <- head(myData)
dataset <- subsetdata[1:7]
knitr::kable(dataset) %>%
  row_spec(0, bold = TRUE) %>%
  column_spec(1, bold = TRUE)

dataset <- subsetdata[c(1, 8:13)]
knitr::kable(dataset) %>%
  row_spec(0, bold = TRUE) %>%
  column_spec(1, bold = TRUE)

dataset <- subsetdata[c(1, 14:19)]
knitr::kable(dataset) %>%
  row_spec(0, bold = TRUE) %>%
  column_spec(1, bold = TRUE)

dataset <- subsetdata[c(1, 20:24)]
knitr::kable(dataset) %>%
  row_spec(0, bold = TRUE) %>%
  column_spec(1, bold = TRUE)
```
\newpage

```{r}
data_sum <- summary(myData)
data_sum
```

\newpage
Below there are some histograms which visualizes the distribution of smoking status, age and gender. 

```{r}
ggplot(myData, aes(x=Smoking.Status)) +
  geom_histogram(stat="count", fill = c("pink", "lightblue"), col = "black") +
  ylim(0, 600) +
  labs(caption = "Figure 1: Number of people who are smoking") +
  theme(plot.caption = element_text(size=16)) +
  theme(plot.caption = element_text(size=16, face="italic")) +
  theme(axis.text = element_text(size = 20)) +
  theme(axis.title = element_text(size = 20, face="bold"))

ggplot(myData, aes(x=Gender)) +
  geom_histogram(stat="count", fill = c("pink", "lightblue"), col = "black") +
  ylim(0, 600) +
  labs(caption = "Figure 2: Gender distribution") +
  theme(plot.caption = element_text(size=16)) +
  theme(plot.caption = element_text(size=16, face="italic")) +
  theme(axis.text = element_text(size = 20)) +
  theme(axis.title = element_text(size = 20, face="bold"))

ggplot(myData, aes(x=Age)) +
  geom_histogram(fill = "lightgrey", col = "black") +
  labs(caption = "Figure 3: Age distribution") +
  theme(plot.caption = element_text(size=16, face="italic")) +
  theme(axis.text = element_text(size = 20)) +
  theme(axis.title = element_text(size = 20, face="bold"))

```


### Plotting the data
These are all the CpG sites plotted against age, with smokin status as color groups and gender as shape groups. A few of the CpG islands have been removed, becaused they looked very similar. What stands out here is that you see two groups in almost every graph, one of men and one of women, so apparently the cg methylation rate is different between men and women.

```{r, results = FALSE}
plotfunction <- function(cg){
  ggplot(data = myData, mapping = aes_string(x = "Age", y = cg)) +
    geom_point(aes(shape = factor(Gender), color = factor(Smoking.Status))) +
    labs(caption = paste(cg, ": Scatterplot visualizing the methylation rate on this CpG island of different ages and genders.")) +
    theme(plot.caption = element_text(size=14, face="italic")) +
    theme(axis.text = element_text(size = 20)) +
    theme(axis.title = element_text(size = 20, face="bold")) +
    theme(legend.text = element_text(size = 16)) +
    theme(legend.title = element_text(size=16)) +
    ylab(paste("CpG Island: ", cg))
}
lapply(names(myData[c(5, 6, 7, 9, 10, 15, 20)]), plotfunction)

```


With the age distribution in mind, let's try to plot 2 methylation rates with the age factored as groups.

```{r}

age.groups <- cut(myData$Age, breaks = c(20, 35, 55, 80), labels = c("20-35y", "36-55y", "55-80y"))

myData$ClassAge <- factor(age.groups)
ggplot(data = myData, mapping = aes(x = ClassAge, y = cg00050873, fill = Smoking.Status)) +
    geom_boxplot() +
    labs(caption = paste("cg00050873: Scatterplot visualizing the methylation rate on this CpG island of different ages and genders.")) +
    theme(plot.caption = element_text(size=14, face="italic")) +
    theme(axis.text = element_text(size = 20)) +
    theme(axis.title = element_text(size = 20, face="bold")) +
    theme(legend.text = element_text(size = 16)) +
    theme(legend.title = element_text(size=16)) +
    ylab(paste("CpG Island: cg00050873")) +
    xlab("Age")

ggplot(data = myData, mapping = aes(x = ClassAge, y = cg03706273, fill = Smoking.Status)) +
    geom_boxplot() +
    labs(caption = paste("cg03706273: Scatterplot visualizing the methylation rate on this CpG island of different ages and genders.")) +
    theme(plot.caption = element_text(size=14, face="italic")) +
    theme(axis.text = element_text(size = 20)) +
    theme(axis.title = element_text(size = 20, face="bold")) +
    theme(legend.text = element_text(size = 16)) +
    theme(legend.title = element_text(size=16)) +
    ylab(paste("CpG Island: cg03706273")) +
    xlab("Age")

```

```{r, echo = FALSE}
myData.pca <- prcomp(myData[,c(5:24)], center = TRUE, scale. = FALSE)
summary(myData.pca)
ggbiplot(myData.pca, ellipse = TRUE, circle = TRUE, scale = 0.5, repel = TRUE, groups = myData$Smoking.Status) + 
  labs(caption = "Figure 4: PCA graph displaying smoking status as different groups") +
  theme(plot.caption = element_text(size=16, face="italic")) +
  theme(axis.text = element_text(size = 20)) +
  theme(axis.title = element_text(size = 20, face="bold")) +
    theme(legend.text = element_text(size = 16)) +
    theme(legend.title = element_text(size=16))

ggbiplot(myData.pca, ellipse = TRUE, circle = TRUE, scale = 0.5, groups = myData$Gender) + 
  labs(caption = "Figure 5: PCA graph displaying gender as different groups") +
  theme(plot.caption = element_text(size=16, face="italic")) +
  theme(axis.text = element_text(size = 20)) +
  theme(axis.title = element_text(size = 20, face="bold")) +
    theme(legend.text = element_text(size = 16)) +
    theme(legend.title = element_text(size=16))
```
